% ****************************************************************************************
%              Title
% ****************************************************************************************

\frenchspacing
\raggedbottom

\pagenumbering{roman}
\pagestyle{plain}

\input{title}

\chapter*{abstract}
The design space for JIT compilers is big, and Julia represents one viewpoint.
The outstanding features of this viewpoint is simplicity and efficiency, which
are enabled by a clever co-design of the language and its implementation.
The combination of simplicity and efficiency also allows users to employ
language strengths and avoid common pitfalls that threats the wide family of JIT
compilers.

My work has been focused on type stability in Julia---a program property
enabling key optimizations in the compiler.
Informally, a function is type stable if the type of
the output depends only on the types of the inputs, not their values.
I have been working on several
challenges related to type stability.
First, gathering data on how widespread the
property is in publicly available Julia code, and what features may be related
to the property.
Second, capturing a formal model of a JIT recognizing the
property at run time and performing optimizations accordingly. Third, still in
works, designing an automated approach to approximate type stability without
running the compiler. Based on this approach, I wish to allow programmers expressing
the assumptions about type stability directly in the code.

%*******************************************************
% Table of Contents (uncomment \tableofcontents)
%*******************************************************
\pagestyle{scrheadings}
%\phantomsection
\pdfbookmark[1]{\contentsname}{tableofcontents}
\setcounter{tocdepth}{2} % <-- 2 includes up to subsections in the ToC
\setcounter{secnumdepth}{3} % <-- 3 numbers up to subsubsections
\tableofcontents

%********************************************************************
% Mainmatter
%*******************************************************
\cleardoublepage
\pagestyle{scrheadings}
\pagenumbering{arabic}
%\setcounter{page}{90}
% use \cleardoublepage here to avoid problems with pdfbookmark
\cleardoublepage
